---
title: 'EDS 223: assignment 4'
author: "Colleen McCamy"
date: "2022-11-16"
output:
    html_document:
      print_df: paged
      toc: yes
      toc_depth: 4
      toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
getwd()
```

## Overview

Marine aquaculture has the potential to play an important role in the global food supply as a more sustainable protein option than land-based meat production.[^1] [Gentry et al.](https://www.nature.com/articles/s41559-017-0257-9) mapped the potential for marine aquaculture globally based on multiple constraints, including ship traffic, dissolved oxygen, bottom depth .[^2]

[^1]: Hall, S. J., Delaporte, A., Phillips, M. J., Beveridge, M. & O'Keefe, M. Blue Frontiers: Managing the Environmental Costs of Aquaculture (The WorldFish Center, Penang, Malaysia, 2011).

[^2]: Gentry, R. R., Froehlich, H. E., Grimm, D., Kareiva, P., Parke, M., Rust, M., Gaines, S. D., & Halpern, B. S. Mapping the global potential for marine aquaculture. *Nature Ecology & Evolution*, 1, 1317-1324 (2017).

For this assignment, you are tasked with determining which Exclusive Economic Zones (EEZ) on the West Coast of the US are best suited to developing marine aquaculture for several species of oysters.\

Based on previous research, we know that oysters needs the following conditions for optimal growth:\

-   sea surface temperature: 11-30°C\
-   depth: 0-70 meters below sea level

##### Learning objectives:

-   combining vector/raster data\
-   resampling raster data\
-   masking raster data\
-   map algebra\

### Data

#### Sea Surface Temperature

We will use average annual sea surface temperature (SST) from the years 2008 to 2012 to characterize the average sea surface temperature within the region. The data we are working with was originally generated from [NOAA's 5km Daily Global Satellite Sea Surface Temperature Anomaly v3.1](https://coralreefwatch.noaa.gov/product/5km/index_5km_ssta.php).

#### Bathymetry

To characterize the depth of the ocean we will use the [General Bathymetric Chart of the Oceans (GEBCO)](https://www.gebco.net/data_and_products/gridded_bathymetry_data/#area).[^3]

[^3]: GEBCO Compilation Group (2022) GEBCO_2022 Grid (<doi:10.5285/e0f0bb80-ab44-2739-e053-6c86abc0289c>).

#### Exclusive Economic Zones

We will be designating maritime boundaries using Exclusive Economic Zones off of the west coast of US from [Marineregions.org](https://www.marineregions.org/eez.php).

## Assignment

Below is an outline of the steps you should consider taking to achieve the assignment tasks.

#### Prepare data (5 points)

To start, we need to load all necessary data and make sure it has the coordinate reference system.

-   load necessary packages and set path 
    -   I recommend using the [`here` package](https://here.r-lib.org/)
-   read in the shapefile for the West Coast EEZ (`wc_regions_clean.shp`)\
-   read in SST rasters
    -   `average_annual_sst_2008.tif`\
    -   `average_annual_sst_2009.tif`\
    -   `average_annual_sst_2010.tif`\
    -   `average_annual_sst_2011.tif`\
    -   `average_annual_sst_2012.tif`\
-   combine SST rasters into a raster stack\
-   read in bathymetry raster (`depth.tif`)\
-   check that data are in the same coordinate reference system\
    -   reproject any data not in the same projection\

```{r include=TRUE, warning=FALSE, messages=FALSE}

# loading in the libraries
library(tidyverse)
library(sf)
library(stars)
library(tmap)
library(raster)
library(terra)
library(tmap)
library(ggplot2)
library(here)
library(OpenStreetMap)
library(tmaptools)

```

```{r}

# reading in the sea surface temperature data
aa_sst_08 <- rast(here("data", "average_annual_sst_2008.tif"))
aa_sst_09 <- rast(here("data", "average_annual_sst_2009.tif"))
aa_sst_10 <- rast(here("data", "average_annual_sst_2010.tif"))
aa_sst_11 <- rast(here("data", "average_annual_sst_2011.tif"))
aa_sst_12 <- rast(here("data", "average_annual_sst_2012.tif"))

class(aa_sst_08)

# combining the sst data 
list_sst <- list(aa_sst_08, aa_sst_09, aa_sst_10, aa_sst_11, aa_sst_12)
sst_rast <- rast(list_sst)
# adding in a crs
crs(sst_rast)


# plottig to check out the data
tm_shape(sst_rast) +
  tm_raster()

# loading in the bathymetry data
depth_rast <- rast(here("data", "depth.tif"))
crs(depth_rast)

plot(depth_rast)

#bath_sf <- st_as_sf(depth_rast)
print(paste0("The coordinate refence system for both raster data are both WGS 84."))

crs(sst_rast)
crs(depth_rast)

```

#### Process data (10 points)

Next, we need process the SST and depth data so that they can be combined. In this case the SST and depth data have slightly different resolutions, extents, and positions. We don't want to change the underlying depth data, so we will need to resample to match the SST data using the nearest neighbor approach.

-   find the mean SST from 2008-2012\
-   convert SST data from Kelvin to Celsius\
    -   hint: subtract by 273.15\
-   crop depth raster to match the extent of the SST raster\
-   note: the resolutions of the SST and depth data do not match\
    -   resample the NPP data to match the resolution of the SST data using the nearest neighbor approach\
-   check that the depth and SST match in resolution, extent, and coordinate reference system\
    -   hint: can the rasters be stacked?

```{r include=TRUE}

# finding the mean SST from 2008-2012
sst_mean <-  mean(sst_rast)
tm_shape(sst_mean)+
  tm_raster() +
  tm_layout(main.title = "Mean Sea Surface Temperature from 2008-2012")

# calculating the mean from Kelvin to Celsius
sst_mean_c <- sst_mean - 273.15

# cropping depth data to sst raster extent
depth_rast_crop <- crop(depth_rast, sst_mean_c)

# matching the resultion of both rasters
depth_rast_crop_res <- resample(x = depth_rast_crop, 
                               y = sst_mean_c, 
                               method = "near")

#checking the CRS
crs(sst_mean_c) == crs(depth_rast_crop_res)

# stacking the two rasters of depth and sea surface temperature
list_sst_depth <- list(sst_mean_c, depth_rast_crop_res)
sst_depth_rast <- rast(list_sst_depth)

# plotting the new raster
# plot(sst_depth_rast)

```

#### Find suitable locations (20)

In order to find suitable locations for marine aquaculture, we'll need to find locations that are suitable in terms of both SST and depth.

-   reclassify SST and depth data into locations that are suitable for oysters\
    -   hint: set suitable values to `1` and unsuitable values to `NA`\
-   find locations that satisfy both SST and depth conditions\
    -   hint: create an overlay using the `lapp()` function multiplying cell values\

```{r include=TRUE}

# creating a reclassification matrix valid sst locations
rcl_sst <- matrix(c(-Inf, 11, NA, 
                    11, 30, 1,
                    30, Inf, NA),
                  ncol = 3, 
                  byrow = TRUE)

# using the reclassifying matrix to set non-suitable sst to NA
sst_suit <- classify(sst_mean_c, rcl = rcl_sst)

# creating a reclassification matrix valid depth locations
rcl_depth <- matrix(c(-Inf, -70, NA, 
                      -70, 0, 1,
                    0, Inf, NA),
                  ncol = 3, 
                  byrow = TRUE)

# using the reclassifying matrix to set non-suitable depth to NA
depth_suit <- classify(depth_rast_crop_res, rcl = rcl_depth)

# cropping sst and depth data based on the mask
sst_oyster <- crop(sst_mean_c, sst_suit)
depth_oyster <- mask(depth_rast_crop_res, depth_suit)

# reprojecting depth_suit as sst_suit due to error below
depth_suit <- project(depth_suit, crs(sst_suit))

# combining the two rasters
list_oyster <- list(depth_suit, sst_suit)
sst_depth_oyster <- rast(list_oyster)

# overlaying the data
fun_oyster <- function(x, y){return(x * y)}
sst_depth_suitable <- lapp(sst_depth_oyster,
                               fun = fun_oyster)

#sst_depth_oyster_apply <- sst_depth_oyster$depth * sst_depth_oyster$mean
tm_shape(sst_depth_suitable) +
  tm_raster() +
  tm_layout("Suitable Locations for Oysters based on Sea Surface Temp. and Depth")

```

#### Determine the most suitable EEZ (20 points)

We want to determine the total suitable area within each EEZ in order to rank zones by priority. To do so, we need to find the total area of suitable locations within each EEZ.

-   select suitable cells within West Coast EEZs\
-   find area of grid cells
-   find the total suitable area within each EEZ\
    -   hint: it might be helpful to rasterize the EEZ data\
-   find the percentage of each zone that is suitable\
    -   hint it might be helpful to join the suitable area by region onto the EEZ vector data\

```{r include=TRUE}

#loading in the data and plotting to check it out
wc_eez_vect <- vect(here("data", "wc_regions_clean.shp"))
# plot(wc_eez_vect)
# wc_eez_vect

wc_eez_vect_area <- shapefile(here("data",
                              "wc_regions_clean.shp"))

wc_eez_vect_area$Area <- area(wc_eez_vect_area)

#print(wc_eez_vect_area$Area)

wc_eez_df <- as_tibble(wc_eez_vect_area) |> 
  dplyr::select(rgn_key, area_km2)
print(paste("The area of the grid cells are", 
            wc_eez_df[1, 1, 1],  "with an area of", 
            round(wc_eez_df[1, 2, 1]), "km^2",
            wc_eez_df[2, 1, 2],  "with an area of",
            round(wc_eez_df[2,2,1]), "km^2",
             wc_eez_df[3, 1, 2],  "with an area of",
            round(wc_eez_df[3,2,1]), "km^2",
            wc_eez_df[4, 1, 2],  "with an area of",
            round(wc_eez_df[4,2,1]), "km^2",
            "."))

# converting to a raster
nams <- names(wc_eez_vect)

wc_eez_rast <- lapply(nams, function(x) {
  rasterize(wc_eez_vect, sst_depth_suitable,
    field = x,
    touches = TRUE)})

# merging all objects into one raster
wc_eez_rast <- do.call("c", wc_eez_rast)

#plot(wc_eez_rast)

# using the suitable locations to mask the wc_eez_rast
wc_eez_rast <- project(wc_eez_rast, 
                       sst_depth_suitable)

# finding the area of a raster cell
cell_area <- cellSize(wc_eez_rast, unit = "km")

# plotting cell size
tm_shape(cell_area) +
  tm_raster() +
  tm_layout(main.title = "Plotting Cell Size")


# masking the suitable areas with the wc_eez_raster
wc_eez_suitable <- mask(wc_eez_rast$rgn, 
                             sst_depth_suitable)

#plot(wc_eez_suitable)

# extracting the areas
area_suitable <- expanse(wc_eez_suitable, unit = "km")
print(paste0("The total suitable area for oysters based on sea surface temperature and depth is ", round(area_suitable), " km^2."))

# extracting area per suitable region and turning it into a dataframe
area_region <- expanse(wc_eez_suitable, unit = "km", byValue = TRUE)
area_region <- as_tibble(area_region)

# adding a region column in the dataframe
region <- tribble(
  ~region, ~rgn_key, 
  "Central California", "CA-C",
  "Northern California", "CA-N",
  "Oregon", "OR",
  "Southern California", "CA-S",
  "Washington", "WA")

area_region <- cbind(area_region, region)

# printing the answers
print(paste0("The suitable area for oysters in the ", area_region$region[1], " region is ", round(area_region$area[1]), " km^2."))

print(paste0("The suitable area for oysters in the ", area_region$region[2], " region is ", round(area_region$area[2]), " km^2."))

print(paste0("The suitable area for oysters in the ", area_region$region[3], " region is ", round(area_region$area[3]), " km^2."))

print(paste0("The suitable area for oysters in the ", area_region$region[4], " region is ", round(area_region$area[4]), " km^2."))

print(paste0("The suitable area for oysters in the ", area_region$region[5], " region is ", round(area_region$area[5]), " km^2."))

# calculating the percent of each area in the suitable region by joining the two dataframes and then calculating the percent per suitable region
area_per <- left_join(area_region, wc_eez_df, by = "rgn_key")
area_per <- area_per |> 
  mutate("area_percent" = area/area_km2 * 100)

# printing the results
print(paste0("The percent of suitable area for oysters in the ", area_per$region[1], " region is about ", round(area_per$area_percent[1], 2), "%."))

print(paste0("The percent of suitable area for oysters in the ", area_per$region[2], " region is about ", round(area_per$area_percent[2], 2), "%."))

print(paste0("The percent of suitable area for oysters in the ", area_per$region[3], " region is about ", round(area_per$area_percent[3], 2), "%."))

print(paste0("The percent of suitable area for oysters in the ", area_per$region[4], " region is about ", round(area_per$area_percent[4], 2), "%."))

print(paste0("The percent of suitable area for oysters in the ", area_per$region[5], " region is about ", round(area_per$area_percent[5], 2), "%."))
```

####  Visualize results (5 points)

Now that we have results, we need to present them!

Create the following maps:

-   total suitable area by region\
-   percent suitable area by region\

Include:

-   legible legends\
-   updated color aesthetics\
-   basemap\

```{r include=TRUE}

# combining area and percent area with the vector and setting it as an sf object
map_data <- merge(wc_eez_vect, area_per, by = "rgn_key")
map_data_sf <- st_as_sf(map_data)
suitable_area <- map_data_sf
percent_suitable_area <- map_data_sf
tmap_mode("view")

# setting the basemap
tmap_style("natural")

# mapping the data with area and percent of area as the fill
tm_shape(suitable_area) +
  tm_fill("area", title = "Total Suitable Area",
              palette= (c("#90e0ef", "#48cae4", "#00b4d8", 
                          "#0096c7", "#0077b6", "#023e8a"))) + tm_borders(col = "#000e30")+
  tm_layout("Suitable Area for Oysters by EEZ Region" ) +
  tm_shape(percent_suitable_area) +
    tm_fill("area_percent", title = "Percent of Area Suitable",
                palette = (c("#92e6a7", "#4ad66d", "#25a244", 
                             "#208b3a","#1a7431","#155d27"))) + tm_borders(col = "#064208")

print("Use layer toggles to switch between total suitable area and the percent of total suitable area per EEZ region.")

```

#### Broaden your workflow! (40 points)

Now that you've worked through the solution for one group of species, let's update your workflow to work for other species. Please create a function that would allow you to reproduce your results for other species. Your function should be able to do the following:\

-   accept temperature and depth ranges and species name as inputs\
-   create maps of total suitable area and percent suitable area per EEZ with the species name in the title\

Run your function for a species of your choice! You can find information on species depth and temperature requirements on [SeaLifeBase](https://www.sealifebase.ca/search.php). Remember, we are thinking about the potential for marine aquaculture, so these species should have some reasonable potential for commercial consumption.

<https://www.sealifebase.ca/summary/Patiria-miniata.html>

```{r}

print("In order to use this funciton make sure you have read in and processed the data in the first part of this document.")

fun_suitable_range <- function(temp_min, temp_max, depth_min, depth_max, species){
  rcl_sst_sp <- matrix(c(-Inf, temp_min, NA, 
                         temp_min, temp_max, 1,
                         temp_max, Inf, NA),
                       ncol = 3, 
                       byrow = TRUE)
  
  sst_suit_sp <- classify(sst_mean_c, rcl_sst_sp)
  
    
  rcl_depth_sp <- matrix(c(-Inf, depth_max, NA,
                           depth_max, depth_min, 1,
                           depth_min, Inf, NA),
                         ncol = 3, 
                         byrow = TRUE)
    
  depth_suit_sp <- classify(depth_rast_crop_res, 
                         rcl = rcl_depth_sp)
    
  sst_crop_sp <- crop(sst_mean_c, sst_suit)
  depth_mask_sp <- mask(depth_rast_crop_res, depth_suit_sp)
  depth_suit_sp <- project(depth_suit_sp, crs(sst_suit_sp))
  list_suit_sp <- list(depth_suit_sp, sst_suit_sp)
  sst_depth_suit_sp <- rast(list_suit_sp)
  fun_suitable_sp <- function(x, y){return(x * y)}
  sst_depth_suitable_sp <- lapp(sst_depth_suit_sp,
                               fun = fun_suitable_sp)
  wc_eez_suitable_sp <- mask(wc_eez_rast$rgn, 
                            sst_depth_suitable_sp)
  area_suitable_sp <- expanse(wc_eez_suitable_sp, unit = "km")
  print(paste0("The total suitable area for ", species, " based on sea surface temperature and depth is ",
                 round(area_suitable_sp), " km^2."))
  area_region_sp <- expanse(wc_eez_suitable_sp, unit = "km", byValue = TRUE)
  area_region_sp <- as_tibble(area_region_sp)
  area_region_sp <- cbind(area_region_sp, region)
  print(paste0("The suitable area for ",  species,  " in the ", area_region_sp$region[1], " region is ",
               round(area_region_sp$area[1]), " km^2."))
  print(paste0("The suitable area for ", species, " in the ", area_region_sp$region[2], " region is ", round(area_region_sp$area[2]), " km^2."))
  print(paste0("The suitable area for ", species, " in the ", area_region_sp$region[3], " region is ",
                 round(area_region_sp$area[3]), " km^2."))
  print(paste0("The suitable area for ",species, "  in the ", area_region_sp$region[4], " region is ",
                 round(area_region_sp$area[4]), " km^2."))
  print(paste0("The suitable area for ", species, " in the ", area_region$region[5], " region is ",
                 round(area_region$area[5]), " km^2."))
    
  area_per_sp <- left_join(area_region_sp, wc_eez_df, by = "rgn_key")
  area_per_sp <- area_per_sp |> 
    mutate("area_percent" = area/area_km2 * 100)
    
  print(paste0("The suitable area for ",  species,  " in the ", area_region_sp$region[1], " region is ",
                 round(area_per_sp$area_percent[1], 2), "%"))
  print(paste0("The suitable area for ",  species,  " in the ", area_region_sp$region[2], " region is ",
                 round(area_per_sp$area_percent[2], 2), "%"))
  print(paste0("The suitable area for ",  species,  " in the ", area_region_sp$region[3], " region is ",
                 round(area_per_sp$area_percent[3], 2), "%"))
  print(paste0("The suitable area for ",  species,  " in the ", area_region_sp$region[4], " region is ",
                 round(area_per_sp$area_percent[4], 2), "%"))
  print(paste0("The suitable area for ",  species,  " in the ", area_region_sp$region[5], " region is ",
                 round(area_per_sp$area_percent[5], 2), "%"))
    
  map_data_sp <- merge(wc_eez_vect, area_per_sp, by = "rgn_key")
  map_data_sp_sf <- st_as_sf(map_data_sp)
  
  suitable_area_sp <- map_data_sp_sf
  percent_suitable_area_sp <- map_data_sp_sf
  
  tmap_mode("view")
  tmap_style("natural")
  
  print(paste0("Suitable Area for ", species, " by EEZ Region"))
  
  tm_shape(suitable_area_sp) +
    tm_fill("area", title = "Total Suitable Area",
            palette = (c("#90e0ef", "#48cae4", "#00b4d8",
                         "#0096c7", "#0077b6", "#023e8a"))) + tm_borders(col = "#000e30")+
     tm_shape(percent_suitable_area_sp) +
       tm_fill("area_percent", title = "Percent of Area Suitable",
               palette = (c("#92e6a7", "#4ad66d", "#25a244",
                            "#208b3a", "#1a7431", "#155d27")))+ tm_borders(col = "#064208")
  #print("Use layer toggles to switch between total suitable area and the percent of total suitable area per EEZ region.")
}

# The bat star has a sea surface temperature range of approximately 2-20 degrees (C) and can live in depths from 0-302 feet below sea level. They are an interesting species to look at as they are affected by sea star wasting disease which is correlated to an increase in ocean temperatures.

fun_suitable_range(temp_min = 2, temp_max = 20, 
                   depth_min = 0, depth_max = -302, 
                   species = "Bat Star")

```
